---
title: "Linear Regression Using Top 6 Principal Components"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Load Libraries

```{r}
library(ggplot2)
library(dplyr)
library(car)        # for VIF
library(caret)      # for RMSE and data splitting
```

## 2. Load and Filter the Data

```{r}
# Load cleaned Boston dataset
data <- read.csv("../data/boston_cleaned.csv")

# Check structure
str(data)
```

## 3. Principal Component Analysis (PCA)

```{r}
# Run PCA on standardized predictors
pr.out <- prcomp(data %>% select(-MEDV), scale. = TRUE)

# Extract top 6 principal components
pc_data <- as.data.frame(pr.out$x[, 1:6])
pc_data$MEDV <- data$MEDV
```

## 4. Split Data (80/20)

```{r}
# Set seed for reproducibility
set.seed(2641)

# Split into training (80%) and testing (20%)
splitIndex <- createDataPartition(pc_data$MEDV, p = 0.8, list = FALSE)
train <- pc_data[splitIndex, ]
test  <- pc_data[-splitIndex, ]
```

## 5. Fit Linear Regression Model

```{r}
# Fit model on training data
PCReg <- lm(MEDV ~ ., data = train)
summary(PCReg)
```

## 6. Predict & Evaluate on Test Set

```{r}
# Predict on test set
predictions <- predict(PCReg, newdata = test)

# Calculate SSE and SST
SSE <- sum((test$MEDV - predictions)^2)
ybar <- mean(train$MEDV)
SST <- sum((test$MEDV - ybar)^2)

# Out-of-sample R-squared
R2_out <- 1 - SSE/SST
cat("Out-of-sample R²:", round(R2_out, 4))
```

## 7. Residual Plot

```{r}
# Residual plot
plot(PCReg$residuals, main = "Residual Plot", ylab = "Residuals", xlab = "Index")
abline(h = 0, col = "red")
```

## 8. Model Validation

### a. T-tests and F-test (from `summary()`)

```{r}
summary(PCReg)  # includes t-values and F-statistic
```

### b. VIF (Variance Inflation Factor)

```{r}
vif(PCReg)  # check multicollinearity
```

---

## Summary

- The model uses **PC1–PC6** which together explain ~93–94% of variance in the predictors.
- The **out-of-sample R²** is `r round(R2_out, 4)`, showing how well the model generalizes.
- **No multicollinearity** is expected since PCs are orthogonal (VIF ~ 1).
- Model significance assessed via **t-tests** and **F-test**.
