---
title: "Boston Housing Data - Full Exploratory Visualization"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set Working Directory

```{r}
setwd("/Users/Ian/Documents/Code_Projects/boston-house-price-modeling/visualization")
getwd()
```

## Load Libraries

```{r}
library(ggplot2)
library(dplyr)
library(corrplot)
library(patchwork)
```

## Load Data

```{r}
data <- read.csv("../data/boston_cleaned.csv")
str(data)
summary(data)
```

## Attribute Information

``` md
Input features in order:
1) CRIM: per capita crime rate by town
2) ZN: proportion of residential land zoned for lots over 25,000 sq.ft.
3) INDUS: proportion of non-retail business acres per town
4) CHAS: Charles River dummy variable (1 if tract bounds river; 0 otherwise)
5) NOX: nitric oxides concentration (parts per 10 million) [parts/10M]
6) RM: average number of rooms per dwelling
7) AGE: proportion of owner-occupied units built prior to 1940
8) DIS: weighted distances to five Boston employment centres
9) RAD: index of accessibility to radial highways
10) TAX: full-value property-tax rate per $10,000 [$/10k]
11) PTRATIO: pupil-teacher ratio by town
12) LSTAT: % lower status of the population

Output variable:
1) MEDV: Median value of owner-occupied homes in $1000's [k$]
```

------------------------------------------------------------------------

## Part 1: Distribution of All Predictive Variables

```{r}
predictor_vars <- names(data)[names(data) != "MEDV"]

# Generate histograms for each predictor
for (var in predictor_vars) {
  print(
    ggplot(data, aes_string(x = var)) +
      geom_histogram(fill = "skyblue", color = "black", bins = 30) +
      labs(title = paste("Histogram of", var), x = var, y = "Frequency")
  )
}
```

------------------------------------------------------------------------

## Part 2: Relationship Between Predictors and MEDV

```{r}
# Special case for CHAS (categorical)
ggplot(data, aes(x = factor(CHAS), y = MEDV)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Boxplot of MEDV by CHAS (Charles River Dummy)",
       x = "CHAS (0 = No, 1 = Yes)", y = "MEDV")
```

```{r}
# Scatter plots for all other numeric predictors vs MEDV
numeric_vars <- setdiff(predictor_vars, "CHAS")

for (var in numeric_vars) {
  print(
    ggplot(data, aes_string(x = var, y = "MEDV")) +
      geom_point(color = "steelblue") +
      labs(title = paste(var, "vs MEDV"),
           x = var, y = "Median Value (MEDV)")
  )
}
```

------------------------------------------------------------------------

## Correlation Heatmap

```{r}
corr_matrix <- cor(data)
corrplot(corr_matrix, method = "color", type = "upper", tl.cex = 0.8, number.cex = 0.7)
```

------------------------------------------------------------------------

## Composite Plot

### **Part 1: Composite Histogram Plot**

```{r}
# Create individual histograms
hist_plots <- lapply(predictor_vars, function(var) {
  ggplot(data, aes_string(x = var)) +
    geom_histogram(fill = "skyblue", color = "black", bins = 30) +
    #labs(title = paste("Histogram of", var), x = var, y = "") +
    labs(x = var, y = "") +
    theme_minimal(base_size = 10)
})

# Combine using patchwork
combined_hist <- wrap_plots(hist_plots, ncol = 4) +
  plot_annotation(title = "Visualization: Distribution of Predictors")

combined_hist
```

------------------------------------------------------------------------

### **Part 2: Composite Scatter/Box Plot (Predictors vs MEDV)**

```{r}
# Create scatter plots
scatter_plots <- lapply(numeric_vars, function(var) {
  ggplot(data, aes_string(x = var, y = "MEDV")) +
    geom_point(color = "steelblue", alpha = 0.6) +
    labs(title = paste(var, "vs MEDV"), x = var, y = "") +
    theme_minimal(base_size = 10)
})

# Add CHAS boxplot
chas_box <- ggplot(data, aes(x = factor(CHAS), y = MEDV)) +
  geom_boxplot(fill = "orange", alpha = 0.8) +
  labs(title = "CHAS vs MEDV", x = "CHAS (0 = No, 1 = Yes)", y = "") +
  theme_minimal(base_size = 10)

# Combine all
combined_scatter <- wrap_plots(c(scatter_plots, list(chas_box)), ncol = 4) +
  plot_annotation(title = "Visualization: Predictor Relationships with MEDV")

combined_scatter
```

------------------------------------------------------------------------

## Save a high-resolution image

```{r}
ggsave("../output/vis_histograms.png", combined_hist, width = 20, height = 12)
ggsave("../output/vis_scatter.png", combined_scatter, width = 20, height = 12)
```

## Summary

This report provides:

-   Full visualization of all 12 predictors\
-   Clear relationships between each variable and the target (MEDV)\
-   A correlation map to detect multicollinearity
