---
title: "Principal Component Analysis (PCA) on Boston Housing Dataset"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Set Working Directory

```{r}
setwd("/Users/Ian/Documents/Code_Projects/boston-house-price-modeling/descriptive_analysis")
getwd()
```

## 1. Load Libraries

```{r}
library(ggplot2)
library(dplyr)
library(ggrepel)   # for non-overlapping labels
library(patchwork)
```

## 2. Load Dataset

```{r}
# Read the cleaned dataset (assumes 'B' column removed)
data <- read.csv("../data/boston_cleaned.csv")
str(data)
```

## 3. Basic Statistics and Correlation

```{r}
# Correlation matrix
cor(data)
```

## 4. PCA Analysis

```{r}
# Run PCA with scaling
pr.out <- prcomp(data, scale. = TRUE)
```

### 4.1 PCA Object Details

```{r}
# Structure of the PCA object
str(pr.out)

# Means and standard deviations
pr.out$center
pr.out$scale
```

### 4.2 Loadings (Eigenvectors)

```{r}
# Loadings of original variables on PCs
pr.out$rotation
```

### 4.3 Principal Component Scores

```{r}
# Principal component scores (rows = observations, cols = PCs)
head(pr.out$x)
```

## 5. Visualize PCA: Biplot

```{r}
# Basic biplot
biplot(pr.out, scale = 0)
```

```{r}
# Optional: Flip PC1 direction (no impact on interpretation)
pr.out$rotation[,1] = -pr.out$rotation[,1]
pr.out$x[,1] = -pr.out$x[,1]
biplot(pr.out, scale = 0)
```

```{r}
# Extract scores and loadings
scores <- as.data.frame(pr.out$x[, 1:2])  # PC1 and PC2
loadings <- as.data.frame(pr.out$rotation[, 1:2])
loadings$varname <- rownames(loadings)

# Scale loadings for visualization (arrows)
loadings_scaled <- loadings
loadings_scaled[, 1:2] <- loadings_scaled[, 1:2] * 10  # adjust scale for better arrow visibility

# Create biplot
p <- ggplot(scores, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_segment(data = loadings_scaled,
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.2, "cm")),
               color = "darkred", linewidth = 0.8) +
  geom_text_repel(data = loadings_scaled,
                  aes(x = PC1, y = PC2, label = varname),
                  color = "darkred", size = 4) +
  labs(title = "PCA Biplot (PC1 vs PC2)",
       x = paste0("PC1 (", round(summary(pr.out)$importance[2, 1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pr.out)$importance[2, 2] * 100, 1), "%)")) +
  theme_minimal()

# Show plot
print(p)

# Save to file
ggsave(filename = "../output/pca/pca_biplot.png", plot = p, width = 8, height = 8, dpi = 300, bg="white")

```

## 6. Variance Explained

```{r}
# Summary of PCA: includes proportion of variance explained
pca_summary <- summary(pr.out)
pca_summary
```

## 7. Scree Plot & Cumulative Variance

### a. Plot

```{r}
# Extract importance matrix
importance <- pca_summary$importance

# Scree plot and cumulative variance
par(mfrow = c(1, 2))
plot(importance[2,], xlab = "Principal Component",
     ylab = "Proportion of Variance Explained", ylim = c(0, 1),
     type = "b", col = "blue", pch = 19)

plot(importance[3,], xlab = "Principal Component",
     ylab = "Cumulative Proportion of Variance Explained",
     ylim = c(0, 1), type = "b", col = "darkgreen", pch = 19)
par(mfrow = c(1, 1))
```

------------------------------------------------------------------------

### b. ggplot2 Version

```{r}
# Extract variance explained from PCA summary
pca_summary <- summary(pr.out)
importance <- pca_summary$importance

# Create a data frame for plotting
pc_data <- data.frame(
  PC = paste0("PC", 1:ncol(pr.out$rotation)),
  Variance = importance[2, ],
  Cumulative = importance[3, ],
  PC_num = 1:ncol(pr.out$rotation)
)

# Scree Plot
scree_plot <- ggplot(pc_data, aes(x = PC_num, y = Variance)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "steelblue", size = 2) +
  scale_x_continuous(breaks = pc_data$PC_num) +
  labs(title = "Scree Plot",
       x = "Principal Component",
       y = "Proportion Variance Explained") +
  theme_minimal()

# Cumulative Variance Plot
cum_plot <- ggplot(pc_data, aes(x = PC_num, y = Cumulative)) +
  geom_line(color = "darkgreen", size = 1) +
  geom_point(color = "darkgreen", size = 2) +
  scale_x_continuous(breaks = pc_data$PC_num) +
  labs(x = "Principal Component",
       y = "Cumulative Prop. Variance Explained") +
  theme_minimal()

# Combine two plots side-by-side using patchwork
combined_plot <- scree_plot + cum_plot

print(combined_plot)

# Save the combined plot to file
ggsave("../output/pca/pca_scree_and_cumulative_pve.png",
       plot = combined_plot, width = 12, height = 6, dpi = 300)
```

## 8. Interpretation

-   The **loadings** (`rotation`) show how each original variable contributes to the PCs.
-   The **principal component scores** (`x`) are new features that can be used for clustering.
-   From the **variance explained**, you can decide how many components to retain (e.g., enough to capture 90% of the variance).
-   The **biplot** helps visualize the relationship between variables and observations in the reduced PC space.
